---
title: "MSD Open"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rhdf5)
library(dplyr)
```

```{r}
h5ls("MillionSongSubset/A/Y/Y/TRAYYBD128F9335F7A.h5")
```

```{r}
analysis <- h5read("MillionSongSubset/A/Y/Y/TRAYYBD128F9335F7A.h5",
       name = "/analysis")
```

```{r}
metadata <- h5read("MillionSongSubset/A/Y/Y/TRAYYBD128F9335F7A.h5",
                   name = "/metadata")
```

```{r}
musicbrain <- h5read("MillionSongSubset/A/Y/Y/TRAYYBD128F9335F7A.h5",
                   name = "/musicbrainz")
```


```{r}
dim(analysis$segments_pitches)
```

Given that the dimension of factors in `analysis` will widely vary depending on the song, should we try to find a way to project data into some n dimensions or find a way to start working with info for songs with same dimensionality (e.g. only work with values stored under `song` in the `analysis` section).


```{r}
h5ls("MillionSongSubset/B/F/S/TRBFSEX128F9300418.h5")
```
```{r}
md2 <- h5read("MillionSongSubset/B/F/S/TRBFSEX128F9300418.h5",
       name = "metadata")
```

```{r}
analysis2 <- h5read("MillionSongSubset/B/F/S/TRBFSEX128F9300418.h5",
                    name = "/analysis")
```

```{r}
in_folder <- function(folder) {
  lapply(list.files(folder), function(x) {paste(folder, x, sep = "/")})
}
```

```{r}
mss1 <- lapply(in_folder("MillionSongSubset"), in_folder)
```

```{r}
mss2 <- lapply(mss1[[1]], in_folder)
```

```{r}
mss3 <- lapply(mss2[[1]], in_folder)
```

```{r}
mss_files <- unlist(mss3)
```

```{r}
analysis_list <- lapply(mss_files, function(x){
  h5read(x,
         name = "/analysis")
})
```

```{r}
song_info <- data.frame(analysis_list[[1]]$songs)

for(i in 2:length(analysis_list)) {
  new_row <- analysis_list[[i]]$songs
  song_info <- rbind(song_info, new_row)
}
```

Should we leave confidence variables in?

```{r}
song_info <- select(song_info, -c(analysis_sample_rate,
                                  audio_md5,
                                  key,
                                  key_confidence,
                                  mode,
                                  mode_confidence,
                                  time_signature,
                                  time_signature_confidence,
                                  track_id
                                  ))
```

```{r}
song_info_alt <- song_info[, colSums(song_info != 0) > 0]
```

```{r}
dim(song_info_alt)
```

Looks like most of the variables don't contain any information as they are all 0. Since this is in the subset they offer, I am going to guess that this problem persists throughout most if not all of the data set. In order to use this data, we will probably have to find some way to standardize song specific data.