---
title: "ddsPLS Exploration"
author: "Harpeth Lee"
date: "2/4/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(ddsPLS2)
```


```{r, eval=FALSE}
ddsPLS2::ddsPLS2_App()
```

This code chunk opens an applet that can be used to build models using ddsPLS. Note that it requires the $X$ and $Y$ variables as separate csv files.


Code copied from the [simulation_ssdpls2 repository](https://github.com/hlorenzo/simulation_ddspls) created by Hadrien Lorenzo.
```{r, include=FALSE}
get_toy_example <- function(n=50,sqrt_1_minus_sig2=0.95,p=1000){
  # Structure
  repX <- 50
  A <- sqrt_1_minus_sig2*matrix(c(rep(1,repX),rep(0,p-repX)),nrow = 1,byrow = T)
  D <- sqrt_1_minus_sig2*matrix(1,nrow = 1,byrow = T)
  # Observations
  d <- ncol(A)+nrow(A)+ncol(D)
  psi <- MASS::mvrnorm(n = n,mu = rep(0,d),Sigma = diag(d))
  phi <- psi[,1:nrow(A),drop=F]
  epsilonX_info <- psi[,nrow(A)+1:repX]*sqrt(1-sqrt_1_minus_sig2^2)
  epsilonX_noise <- psi[,nrow(A)+repX+1:(ncol(A)-repX)]
  epsilonY_info <- psi[,nrow(A)+ncol(A)+1,drop=F]*sqrt(1-sqrt_1_minus_sig2^2)
  # X and Y
  X <- phi%*%A + cbind(epsilonX_info,epsilonX_noise)
  Y <- phi%*%D + epsilonY_info
  list(X=X,Y=Y)
}
```


```{r}
# Creates a toy data set for the ddsPLS function
toy_ex <- get_toy_example()
```

```{r}
# Creates model from the toy data
toy_mod <- ddsPLS(toy_ex$X, toy_ex$Y)

toy_results <- toy_mod$results
```


## Recreate Toy Example

This is a recreation of the toy example created by Hadrien Lorenzo, the original example can be found [here](http://htmlpreview.github.io/?https://github.com/hlorenzo/simulation_ddspls/blob/main/simulation_ddspls.html).

```{r, message=TRUE}
# Creates toy data set to be used
simu_toy <- get_toy_example(n=50,sqrt_1_minus_sig2 = 0.9025,p = 1000)

# Creates vector of lambda values to be used
lambdas <- seq(0,1,length.out = 30)

# Sets number of bootstrap samples to run
n_B <- 100

# Creates model using ddsPLS algorithm
model_toy <- ddsPLS(simu_toy$X,simu_toy$Y,
                    doBoot = FALSE,
                   lambdas = lambdas,
                   n_B = n_B,
                   verbose = T # whether trace during process
                   )
```

```{r}
model_toy_2 <- ddsPLS(simu_toy$X,simu_toy$Y,
                      doBoot = FALSE,
                  criterion = "Q2",
                   lambdas = lambdas,
                   n_B = n_B,
                   verbose = T # whether trace during process
                   )
```


## Design 1

```{r, include=FALSE}
get_design_1 <- function(n=50,sqrt_1_minus_sig2=0.99,p=1000,q=3){
  # Structure
  alpha3 <- 1/sqrt(3)
  alpha2 <- 1/sqrt(2)
  repX <- 50
  A1 <- c(rep(alpha3,repX),rep(0,p-repX))
  A2 <- c(rep(0,repX),rep(alpha2,repX),rep(0,p-2*repX))
  A <- matrix(c(rep(A1,3),rep(A2,2)),nrow = 5,byrow = T)*sqrt_1_minus_sig2
  D1 <- c(rep(alpha3,1),rep(0,q-1))
  D2 <- c(rep(0,1),rep(alpha2,1),rep(0,q-2))
  D <- matrix(c(rep(D1,3),rep(D2,2)),nrow = 5,byrow = T)*sqrt_1_minus_sig2
  # Observations
  d <- ncol(A)+nrow(A)+ncol(D)
  psi <- MASS::mvrnorm(n = n,mu = rep(0,d),Sigma = diag(d))
  phi <- psi[,1:nrow(A)]
  epsilonX_info <- psi[,nrow(A)+1:(2*repX)]*sqrt(1-sqrt_1_minus_sig2^2)
  epsilonX_noise <- psi[,nrow(A)+(2*repX)+1:(ncol(A)-2*repX)]
  epsilonY_info <- psi[,nrow(A)+ncol(A)+1:2]*sqrt(1-sqrt_1_minus_sig2^2)
  epsilonY_noise <- psi[,d]
  # X and Y
  X <- phi%*%A + cbind(epsilonX_info,epsilonX_noise)
  Y <- phi%*%D + cbind(epsilonY_info,epsilonY_noise)
  list(X=X,Y=Y)
}
```



```{r}
simu_1 <- get_design_1(n=50,sqrt_1_minus_sig2 = 0.99,p = 1000,q = 3)
```

What does the `NCORES` argument do? Setting it to integers greater than 1 gives an error.

Is there a way to include more components in the model?
```{r}
model_1 <- ddsPLS(simu_1$X,simu_1$Y,
                  lambdas = lambdas,
                  n_B=n_B,
                  verbose=T)
```
```{r}
model_1_Q2 <- ddsPLS(simu_1$X,simu_1$Y,
                  criterion = "Q2",
                  lambdas = lambdas,
                  n_B=n_B,
                  verbose=T)

model_1_lambda <- ddsPLS(simu_1$X, simu_1$Y,
                         criterion = "Q2",
                         lambdas = seq(0,1,length.out = 100),
                         n_B = n_B,
                         verbose = T)

model_1_lambda_2 <- ddsPLS(simu_1$X, simu_1$Y,
                         criterion = "Q2",
                         lambdas = seq(0,1,length.out = 50),
                         n_B = n_B,
                         verbose = T)
```

## Different Simulations of Design 1 Data


There is a problem with `get_design_1`, `q` cannot take values other than 5.
```{r, eval=FALSE}
data_1 <- get_design_1(n = 100, p = 1000, q = 5)

ddsPLS(data_1$X,data_1$Y,
                  criterion = "Q2",
                  lambdas = lambdas,
                  n_B=n_B,
                  verbose=T)
```

## Design 2

```{r, include=FALSE}
get_design_2 <- function(seed=1,n=50,q=5,p1=500,p2=5000,
                         sigma1=0.05,sigma2=0.05,sigmaY=0.1,
                         ncpX=10,ncpXCom=5,ncpXYCom=3,plot=F){
  # Internal Function
  simulateData <- function(n, ncp, p, sigma, sigmaNoise=0.1, ConcVarFact=0.8, n_min_peaks=3){
    meanPeakSigma <- sigma
    sigPeakSigma <- sigma / 4
    axis <- 1:p
    S <- matrix(0,ncp,p)
    C <- matrix(0,n,ncp)
    for(i in 1:ncp){
      npeaks <- 3+ceiling(10*runif(1))
      peakheights <- runif(npeaks)
      sigmas <- runif(npeaks) * sigPeakSigma + meanPeakSigma
      position <- runif(npeaks) * p
      for(j in 1:npeaks){
        S[i,] <- S[i,] + peakheights[j] * exp(-0.5 * ((axis - position[j]) / sigmas[j])^2)
      }
    }
    meanC <- sort(10^runif(ncp),decreasing = T)
    varC <- ConcVarFact * meanC * runif(ncp)
    for(i in 1:ncp){
      C[,i] <- rnorm(n = n,mean = meanC[i], sd = varC[i]/2)
    }
    X <- C%*%S;X <- X/max(abs(X))
    E <- matrix(rnorm(n*p,sd = sigmaNoise),nrow = n,ncol = p)
    X <- X*sqrt(1-sigmaNoise^2) + E
    list(X=X, C=C, S=S, E=E)
  }

  set.seed(seed)
  # ncpX for each X separately
  Data_1 <- simulateData(n=n, ncp=ncpX, p=p1, sigma=20, sigmaNoise = sigma1, ConcVarFact=0.8, n_min_peaks=5)
  Data_2 <- simulateData(n=n, ncp=ncpX, p=p2, sigma=30, sigmaNoise = sigma2, ConcVarFact=0.8, n_min_peaks=5)
  S1 <- Data_1$S;C1 <- Data_1$C;X1 <- Data_1$X
  S2 <- Data_2$S;C2 <- Data_2$C
  # ncpXCom in common
  C2[,1:ncpXCom] <- C1[,1:ncpXCom,drop=F]
  X2 <- C2%*%S2;X2 <- X2/max(abs(X2))
  E2 <- matrix(rnorm(n*p2,sd = sigma2),nrow = n,ncol = p2)
  X2 <- X2*sqrt(1-sigma2^2) + E2
  # Build Y on ncpXYCom components
  Y <- scale(Data_1$C[,1:ncpXYCom,drop=F])
  # Add extra variables and noise
  E_y <- matrix(rnorm(n*q,sd = sigmaY),nrow = n,ncol = q)
  if(q>ncpXYCom){
    Y <- cbind(Y,matrix(rnorm(n*(q-ncpXYCom)),nrow = n))*sqrt(1-sigmaY^2) + E_y
  }else{
    Y <- Y*sqrt(1-sigmaY^2) + E_y
  }
  if(plot){
    layout(matrix(c(1,2,2,3,3,3),nrow = 2,byrow = T))
    matplot(t(X1),lty=1,type="l")
    matplot(t(X2),lty=1,type="l")
    corXY <- cor(cbind(X1,X2),Y)
    matplot(corXY,type="l")
  }
  list(Xs=list(X1=X1,X2=X2),Y=Y,S=list(S1=S1,S2=S2,SXY=list(S1=S1[1:ncpXYCom,],S2=S2[1:ncpXYCom,])))
}
```

```{r}
simu_2.1 <- get_design_2(plot = T)
```

```{r}
model_2.1 <- ddsPLS(simu_2.1$X$X1, simu_2.1$Y,
                    criterion = "Q2",
                    n_lambdas = 25,
                    verbose = TRUE)

model_2.2 <- ddsPLS(simu_2.1$X$X2, simu_2.1$Y,
                    criterion = "Q2",
                    n_lambdas = 25,
                    verbose = TRUE)
```

```{r}
simu_2.2 <- get_design_2(seed = 2, ncpX = 20, plot = T)
```

```{r}
model_2.3 <- ddsPLS(simu_2.2$Xs$X1, simu_2.2$Y,
                    criterion = "Q2",
                    n_lambdas = 30,
                    verbose = TRUE)

model_2.4 <- ddsPLS(simu_2.2$Xs$X2, simu_2.2$Y,
                    criterion = "Q2",
                    n_lambdas = 30,
                    verbose = TRUE)
```

Model results can also be plotted using the `plot` function.
```{r}
plot(model_2.1,type="Q2",legend.position = "bottomleft")
```


## Get Data Simulation

The following `get_data` function is from the vignette for the `ddsPLS` package
```{r, include=FALSE}
get_data <- function(n=100,alpha=0.4,beta_0=0.2,eps=0.95,
                     p1=50,p2=25,p3=25,p=1000){
   R1 = R2 = R3 <- 1
   d <- R1+R2+R3
   A0 <- matrix(c(
     c(rep(1/sqrt(R1),p1),rep(sqrt(alpha),p2),rep(0,p3),rep(0,p-p1-p2-p3)),
     c(rep(0,p1),rep(sqrt(1-alpha),p2),rep(0,p3),rep(0,p-p1-p2-p3)),
     c(rep(0,p1),rep(0,p2),rep(1,p3),rep(0,p-p1-p2-p3))
   ),nrow = d,byrow = T)
   A <- eps*A0
   D0 <- matrix(c(1,0,0,
                  sqrt(beta_0),sqrt(1-beta_0),0,
                  0,0,0),nrow = d,byrow = F)
   D <- eps*D0
   q <- ncol(D)
   L_total <- q+p
   psi <- MASS::mvrnorm(n,mu = rep(0,d+L_total),Sigma = diag(d+L_total))
   phi <- psi[,1:d,drop=F]
   errorX <- matrix(rep(sqrt(1-apply(A^2,2,sum)),n),n,byrow = T)
   errorY <- matrix(rep(sqrt(1-apply(D^2,2,sum)),n),n,byrow = T)
   X <- phi%*%A + errorX*psi[,d+1:p,drop=F]
   Y <- phi%*%D + errorY*psi[,d+p+1:q,drop=F]
   list(X=X,Y=Y)
 }
```

```{r}
data_3.1 <- get_data()

model_3.1 <- ddsPLS(data_3.1$X, data_3.1$Y,
                    criterion = "Q2",
                    n_lambdas = 30,
                    verbose = TRUE)
```

```{r}
data_3.2 <- get_data(p1 = 50, p2 = 50, p3 = 50, p = 250)

model_3.2 <- ddsPLS(data_3.2$X, data_3.2$Y,
                    criterion = "Q2",
                    n_lambdas = 30,
                    verbose = TRUE)

model_3.3 <- ddsPLS(data_3.2$X, data_3.2$Y,
                    criterion = "Q2",
                    n_lambdas = 30,
                    verbose = TRUE,
                    LD = TRUE)

model_3.4 <- ddsPLS(data_3.2$X, data_3.2$Y,
                    criterion = "diffR2Q2",
                    n_lambdas = 30,
                    verbose = TRUE,
                    LD = TRUE)
```


