---
title: "ddsPLS Exploration"
author: "Harpeth Lee"
date: "2/4/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(ddsPLS2)
library(MASS)
library(spls)
```


```{r, eval=FALSE}
ddsPLS2::ddsPLS2_App()
```

This code chunk opens an applet that can be used to build models using ddsPLS. Note that it requires the $X$ and $Y$ variables as separate csv files.


Code copied from the [simulation_ssdpls2 repository](https://github.com/hlorenzo/simulation_ddspls) created by Hadrien Lorenzo.
```{r Toy Data Simulation, include=FALSE}
get_toy_example <- function(n=50,sqrt_1_minus_sig2=0.95,p=1000){
  # Structure
  repX <- 50
  A <- sqrt_1_minus_sig2*matrix(c(rep(1,repX),rep(0,p-repX)),nrow = 1,byrow = T)
  D <- sqrt_1_minus_sig2*matrix(1,nrow = 1,byrow = T)
  # Observations
  d <- ncol(A)+nrow(A)+ncol(D)
  psi <- MASS::mvrnorm(n = n,mu = rep(0,d),Sigma = diag(d))
  phi <- psi[,1:nrow(A),drop=F]
  epsilonX_info <- psi[,nrow(A)+1:repX]*sqrt(1-sqrt_1_minus_sig2^2)
  epsilonX_noise <- psi[,nrow(A)+repX+1:(ncol(A)-repX)]
  epsilonY_info <- psi[,nrow(A)+ncol(A)+1,drop=F]*sqrt(1-sqrt_1_minus_sig2^2)
  # X and Y
  X <- phi%*%A + cbind(epsilonX_info,epsilonX_noise)
  Y <- phi%*%D + epsilonY_info
  list(X=X,Y=Y)
}
```

The `get_toy_example()` function simulates a data frame with `n` observations of where 50 of `p` predictors are associated with the single response variable.


```{r}
# Creates a toy data set for the ddsPLS function
toy_ex <- get_toy_example()
```

```{r}
# Creates model from the toy data
toy_mod <- ddsPLS(toy_ex$X, toy_ex$Y)

toy_results <- toy_mod$results
```


## Recreate Toy Example

This is a recreation of the toy example created by Hadrien Lorenzo, the original example can be found [here](http://htmlpreview.github.io/?https://github.com/hlorenzo/simulation_ddspls/blob/main/simulation_ddspls.html).

```{r, message=TRUE, cache=TRUE}
# Creates toy data set to be used
simu_toy <- get_toy_example(n=50,sqrt_1_minus_sig2 = 0.9025,p = 1000)

# Creates vector of lambda values to be used
lambdas <- seq(0,1,length.out = 30)

# Sets number of bootstrap samples to run
n_B <- 100

# Creates model using ddsPLS algorithm
model_toy <- ddsPLS(simu_toy$X,simu_toy$Y,
                    doBoot = FALSE,
                   lambdas = lambdas,
                   n_B = n_B,
                   verbose = T # whether trace during process
                   )
```

```{r}
model_toy_2 <- ddsPLS(simu_toy$X,simu_toy$Y,
                      doBoot = FALSE,
                  criterion = "Q2",
                   lambdas = lambdas,
                   n_B = n_B,
                   verbose = T # whether trace during process
                   )
```


## Design 1

```{r Design 1 Simulation, include=FALSE, cache=TRUE}
get_design_1 <- function(n=50,sqrt_1_minus_sig2=0.99,p=1000,q=3){
  # Structure
  alpha3 <- 1/sqrt(3)
  alpha2 <- 1/sqrt(2)
  repX <- 50
  A1 <- c(rep(alpha3,repX),rep(0,p-repX))
  A2 <- c(rep(0,repX),rep(alpha2,repX),rep(0,p-2*repX))
  A <- matrix(c(rep(A1,3),rep(A2,2)),nrow = 5,byrow = T)*sqrt_1_minus_sig2
  D1 <- c(rep(alpha3,1),rep(0,q-1))
  D2 <- c(rep(0,1),rep(alpha2,1),rep(0,q-2))
  D <- matrix(c(rep(D1,3),rep(D2,2)),nrow = 5,byrow = T)*sqrt_1_minus_sig2
  # Observations
  d <- ncol(A)+nrow(A)+ncol(D)
  psi <- MASS::mvrnorm(n = n,mu = rep(0,d),Sigma = diag(d))
  phi <- psi[,1:nrow(A)]
  epsilonX_info <- psi[,nrow(A)+1:(2*repX)]*sqrt(1-sqrt_1_minus_sig2^2)
  epsilonX_noise <- psi[,nrow(A)+(2*repX)+1:(ncol(A)-2*repX)]
  epsilonY_info <- psi[,nrow(A)+ncol(A)+1:2]*sqrt(1-sqrt_1_minus_sig2^2)
  epsilonY_noise <- psi[,d]
  # X and Y
  X <- phi%*%A + cbind(epsilonX_info,epsilonX_noise)
  Y <- phi%*%D + cbind(epsilonY_info,epsilonY_noise)
  list(X=X,Y=Y)
}
```

Generates `n` samples of `p` observations with `q` response variables. Projects 5 latent variables onto `p` components.

```{r}
simu_1 <- get_design_1(n=50,sqrt_1_minus_sig2 = 0.99,p = 1000,q = 3)
```

What does the `NCORES` argument do? Setting it to integers greater than 1 gives an error.

Is there a way to include more components in the model?
```{r, cache=TRUE}
model_1 <- ddsPLS(simu_1$X,simu_1$Y,
                  lambdas = lambdas,
                  n_B=n_B,
                  verbose=T)
```
```{r, cache=TRUE}
model_1_Q2 <- ddsPLS(simu_1$X,simu_1$Y,
                  criterion = "Q2",
                  lambdas = lambdas,
                  n_B=n_B,
                  verbose=T)

model_1_lambda <- ddsPLS(simu_1$X, simu_1$Y,
                         criterion = "Q2",
                         lambdas = seq(0,1,length.out = 100),
                         n_B = n_B,
                         verbose = T)

model_1_lambda_2 <- ddsPLS(simu_1$X, simu_1$Y,
                         criterion = "Q2",
                         lambdas = seq(0,1,length.out = 50),
                         n_B = n_B,
                         verbose = T)
```

## Different Simulations of Design 1 Data


There is a problem with `get_design_1`, `q` cannot take values other than 5.
```{r, eval=FALSE}
data_1 <- get_design_1(n = 100, p = 1000, q = 5)

ddsPLS(data_1$X,data_1$Y,
                  criterion = "Q2",
                  lambdas = lambdas,
                  n_B=n_B,
                  verbose=T)
```

## Design 2

```{r Design 2 Simulation, include=FALSE, cache=TRUE}
get_design_2 <- function(seed=1,n=50,q=5,p1=500,p2=5000,
                         sigma1=0.05,sigma2=0.05,sigmaY=0.1,
                         ncpX=10,ncpXCom=5,ncpXYCom=3,plot=F){
  # Internal Function
  simulateData <- function(n, ncp, p, sigma, sigmaNoise=0.1, ConcVarFact=0.8, n_min_peaks=3){
    meanPeakSigma <- sigma
    sigPeakSigma <- sigma / 4
    axis <- 1:p
    S <- matrix(0,ncp,p)
    C <- matrix(0,n,ncp)
    for(i in 1:ncp){
      npeaks <- 3+ceiling(10*runif(1))
      peakheights <- runif(npeaks)
      sigmas <- runif(npeaks) * sigPeakSigma + meanPeakSigma
      position <- runif(npeaks) * p
      for(j in 1:npeaks){
        S[i,] <- S[i,] + peakheights[j] * exp(-0.5 * ((axis - position[j]) / sigmas[j])^2)
      }
    }
    meanC <- sort(10^runif(ncp),decreasing = T)
    varC <- ConcVarFact * meanC * runif(ncp)
    for(i in 1:ncp){
      C[,i] <- rnorm(n = n,mean = meanC[i], sd = varC[i]/2)
    }
    X <- C%*%S;X <- X/max(abs(X))
    E <- matrix(rnorm(n*p,sd = sigmaNoise),nrow = n,ncol = p)
    X <- X*sqrt(1-sigmaNoise^2) + E
    list(X=X, C=C, S=S, E=E)
  }

  set.seed(seed)
  # ncpX for each X separately
  Data_1 <- simulateData(n=n, ncp=ncpX, p=p1, sigma=20, sigmaNoise = sigma1, ConcVarFact=0.8, n_min_peaks=5)
  Data_2 <- simulateData(n=n, ncp=ncpX, p=p2, sigma=30, sigmaNoise = sigma2, ConcVarFact=0.8, n_min_peaks=5)
  S1 <- Data_1$S;C1 <- Data_1$C;X1 <- Data_1$X
  S2 <- Data_2$S;C2 <- Data_2$C
  # ncpXCom in common
  C2[,1:ncpXCom] <- C1[,1:ncpXCom,drop=F]
  X2 <- C2%*%S2;X2 <- X2/max(abs(X2))
  E2 <- matrix(rnorm(n*p2,sd = sigma2),nrow = n,ncol = p2)
  X2 <- X2*sqrt(1-sigma2^2) + E2
  # Build Y on ncpXYCom components
  Y <- scale(Data_1$C[,1:ncpXYCom,drop=F])
  # Add extra variables and noise
  E_y <- matrix(rnorm(n*q,sd = sigmaY),nrow = n,ncol = q)
  if(q>ncpXYCom){
    Y <- cbind(Y,matrix(rnorm(n*(q-ncpXYCom)),nrow = n))*sqrt(1-sigmaY^2) + E_y
  }else{
    Y <- Y*sqrt(1-sigmaY^2) + E_y
  }
  if(plot){
    layout(matrix(c(1,2,2,3,3,3),nrow = 2,byrow = T))
    matplot(t(X1),lty=1,type="l")
    matplot(t(X2),lty=1,type="l")
    corXY <- cor(cbind(X1,X2),Y)
    matplot(corXY,type="l")
  }
  list(Xs=list(X1=X1,X2=X2),Y=Y,S=list(S1=S1,S2=S2,SXY=list(S1=S1[1:ncpXYCom,],S2=S2[1:ncpXYCom,])))
}
```

```{r, cache=TRUE}
simu_2.1 <- get_design_2(plot = T)
```

```{r, cache=TRUE}
model_2.1 <- ddsPLS(simu_2.1$X$X1, simu_2.1$Y,
                    criterion = "Q2",
                    n_lambdas = 25,
                    verbose = TRUE)

model_2.2 <- ddsPLS(simu_2.1$X$X2, simu_2.1$Y,
                    criterion = "Q2",
                    n_lambdas = 25,
                    verbose = TRUE)
```

```{r, cache=TRUE}
simu_2.2 <- get_design_2(seed = 2, ncpX = 20, plot = T)
```

```{r, cache=TRUE}
model_2.3 <- ddsPLS(simu_2.2$Xs$X1, simu_2.2$Y,
                    criterion = "Q2",
                    n_lambdas = 30,
                    verbose = TRUE)

model_2.4 <- ddsPLS(simu_2.2$Xs$X2, simu_2.2$Y,
                    criterion = "Q2",
                    n_lambdas = 30,
                    verbose = TRUE)
```

Model results can also be plotted using the `plot` function.
```{r, cache=TRUE}
plot(model_2.1,type="Q2",legend.position = "bottomleft")
```


## Get Data Simulation

The following `get_data` function is from the vignette for the `ddsPLS` package
```{r, include=FALSE, cache=TRUE}
get_data <- function(n=100,alpha=0.4,beta_0=0.2,eps=0.95,
                     p1=50,p2=25,p3=25,p=1000){
   R1 = R2 = R3 <- 1
   d <- R1+R2+R3
   A0 <- matrix(c(
     c(rep(1/sqrt(R1),p1),rep(sqrt(alpha),p2),rep(0,p3),rep(0,p-p1-p2-p3)),
     c(rep(0,p1),rep(sqrt(1-alpha),p2),rep(0,p3),rep(0,p-p1-p2-p3)),
     c(rep(0,p1),rep(0,p2),rep(1,p3),rep(0,p-p1-p2-p3))
   ),nrow = d,byrow = T)
   A <- eps*A0
   D0 <- matrix(c(1,0,0,
                  sqrt(beta_0),sqrt(1-beta_0),0,
                  0,0,0),nrow = d,byrow = F)
   D <- eps*D0
   q <- ncol(D)
   L_total <- q+p
   psi <- MASS::mvrnorm(n,mu = rep(0,d+L_total),Sigma = diag(d+L_total))
   phi <- psi[,1:d,drop=F]
   errorX <- matrix(rep(sqrt(1-apply(A^2,2,sum)),n),n,byrow = T)
   errorY <- matrix(rep(sqrt(1-apply(D^2,2,sum)),n),n,byrow = T)
   X <- phi%*%A + errorX*psi[,d+1:p,drop=F]
   Y <- phi%*%D + errorY*psi[,d+p+1:q,drop=F]
   list(X=X,Y=Y)
 }
```

The variable `eps` seems to relate the predictors and response, as well as `phi`. The dimension of `phi` specifies the number of latent variables.

```{r, cache=TRUE}
data_3.1 <- get_data()

model_3.1 <- ddsPLS(data_3.1$X, data_3.1$Y,
                    criterion = "Q2",
                    n_lambdas = 30,
                    verbose = TRUE)
```

```{r, cache=TRUE}
data_3.2 <- get_data(p1 = 50, p2 = 50, p3 = 50, p = 250)

model_3.2 <- ddsPLS(data_3.2$X, data_3.2$Y,
                    criterion = "Q2",
                    n_lambdas = 30,
                    verbose = TRUE)

model_3.3 <- ddsPLS(data_3.2$X, data_3.2$Y,
                    criterion = "Q2",
                    n_lambdas = 30,
                    verbose = TRUE,
                    LD = TRUE)

model_3.4 <- ddsPLS(data_3.2$X, data_3.2$Y,
                    criterion = "diffR2Q2",
                    n_lambdas = 30,
                    verbose = TRUE,
                    LD = TRUE)
```

# Novel Simulations


The following code simulates a data set with 100 uncorrelated predictors and 1 response variable all sampled from a normal distribution.
```{r}
Sigma <- diag(100)

sim_pred <- mvrnorm(n = 1000, mu = rep(0, 100), Sigma = Sigma)

sim_resp <- matrix(rnorm(1000), 1000, 1)
```


```{r}
ddsPLS(sim_pred, sim_resp, verbose = TRUE)
```
As expected no model is built as performance is awful. Interestingly message "no Q2r large enough for tested lambda" is given for justification, seems to suggest it checks just $Q^2$. Perhaps this just means that mean estimation performs better.

```{r}
Sigma <- matrix(c(1,.75,.75,1),2,2)

n <- 20
p <- 5
p <- p - 2

sim_preds <- cbind(mvrnorm(n = n, rep(0, 2), Sigma), matrix(rep(0,n*p),n, p))

sim_resp <- as.matrix(apply(sim_preds,1,function(x) 5*x[1]+x[2]))

sim_preds <- sim_preds + matrix(rnorm(n*(p+2), sd = 0.6), n,(p+2))
sim_resp <- sim_resp + matrix(rnorm(n, sd = 0.8), n,1)
```

The above code simulates data with $n$ observations of $p$ predictors and 1 response variable. There are two predictors from which responses are linearly generated. Random noise is then added to the predictors and the response. As expected, the ddsPLS model performs very well.

```{r}
pls_model <- ddsPLS(sim_preds, sim_resp, verbose = TRUE)
```

## Complex Simulated Data 

The general structure of simulated data is $\textbf{X} = \textbf{A}^T \phi + \epsilon_{X}$ and $\textbf{Y} = \textbf{D}^T \phi + \epsilon_Y$. Note that $\phi$ provides the structure between the two. Code structures it as $\textbf{X} = \phi\textbf{A}  + \epsilon_{X}$ and similarly for $\textbf{Y}$. $\epsilon$ is added random error. $\text{Cov}(\textbf{X}, \textbf{Y}) = \textbf{D}^T \textbf{A}$.


```{r Sim Data Function}
sim_data <- function(n = 5, p = 10, q = 2, R = 5, x = 3, noise_weight = 1, D_method = "new") {
  
  # Ensures x<=R, if x>R the dimension of A is incompatible with phi
  if(x > R){
    x = R
  }
  
  # Creates A and D matrices
  A <- matrix(c(rep(rep(1,p),x), rep(rep(0,p),R-x)), ncol = p)
  
  if(D_method == "new") {
     D <- matrix(rep(1, R*q), nrow = R)
  } else {
    D <- diag(max(q, R))[1:R, 1:q]
  }
  
  d <- ncol(A)+nrow(A)+ncol(D)
  psi <- MASS::mvrnorm(n = n,mu = rep(0,d),Sigma = diag(d))
  phi <- psi[,1:nrow(A)]
  
  epsilon_X <- mvrnorm(n = dim(phi)[1],
                       rep(0, dim(A)[2]),
                       Sigma = noise_weight*diag(dim(A)[2]))
  
  epsilon_Y <- mvrnorm(n = dim(phi)[1],
                       rep(0, dim(D)[2]),
                       Sigma = noise_weight*diag(dim(D)[2]))
  
  X <- phi %*% A + epsilon_X
  Y <- phi %*% D + epsilon_Y
  
  
  list(X=X, Y=Y)
}
```


```{r, eval=FALSE}
# This chunk isn't running, my guess is that setting a seed causes
# problems for the bootstrapping done to run ddsPLS

set.seed(1999)
sim_1 <- sim_data()

sim1_pls <- ddsPLS(sim_1$X, sim_1$Y,
                   n_B = 30,
                   n_lambdas = 25,
                   verbose = TRUE)
```

ddsPLS is not an appropriate method when working with a small sample size as 

```{r, eval=FALSE}
set.seed(1999)
sim_1 <- sim_data()
```

```{r, eval=FALSE}
sim1_pls <- ddsPLS(sim_1$X, sim_1$Y,
                   verbose = TRUE)
```

```{r, cache=TRUE}
set.seed(1894)

sim_2 <- sim_data(n = 100, p = 150, q = 5, R = 5, x = 4, D_method = "old")
```

```{r, cache=TRUE}
sim2_pls <- ddsPLS(sim_2$X, sim_2$Y,
                   verbose = TRUE)

sim2_pls <- ddsPLS(sim_2$X, sim_2$Y,
                   criterion = "Q2",
                   verbose = TRUE)
```

```{r, cache=TRUE}
set.seed(108)

sim_3 <- sim_data(n = 50, p = 200, q = 5, R = 25, x = 10, D_method = "old")
```

```{r, cache=TRUE}
sim3_pls <- ddsPLS(sim_3$X, sim_3$Y,
                   verbose = TRUE)

sim3_pls <- ddsPLS(sim_3$X, sim_3$Y,
                   criterion = "Q2",
                   verbose = TRUE)
```

```{r, cache=TRUE}
set.seed(43)

sim_4 <- sim_data(n = 250, p = 1000, q = 10, R = 100, x = 95, D_method = "old")
```

```{r}
sim4_pls <- ddsPLS(sim_4$X, sim_4$Y,
                   verbose = TRUE)

sim4_pls <- ddsPLS(sim_4$X, sim_4$Y,
                   criterion = "Q2",
                   verbose = TRUE)
```

```{r, cache=TRUE}
set.seed(99)

sim_5 <- sim_data(n = 500, p = 500, q = 10, R = 100, x = 95, D_method = "old")
```

```{r, cache=TRUE}
sim5_pls <- ddsPLS(sim_5$X, sim_5$Y,
                   verbose = TRUE)

sim5_pls <- ddsPLS(sim_5$X, sim_5$Y,
                   criterion = "Q2",
                   verbose = TRUE)
```

```{r,cache=TRUE}
set.seed(99)

sim_6 <- sim_data(n = 250, p = 200, q = 10, R = 8, x = 7, D_method = "old")
```

```{r, cache=TRUE}
sim6_pls <- ddsPLS(sim_6$X, sim_6$Y,
                   verbose = TRUE)

sim6_pls <- ddsPLS(sim_6$X, sim_6$Y,
                   criterion = "Q2",
                   verbose = TRUE)
```

It looks like there is a problem with the way I am simulating the data. All of the simulations have only explained a low amount of the variation in the data, perhaps due to the relation between `R` and `x`.


I changed `A` from a subset of a diagonal matrix to a matrix of all 1s and the function starts to work well. Running the code with different values of `q` and `noise_weight = 0` returns a model that perfectly explains the data.
```{r, cache=TRUE}
set.seed(12)

sim_7 <- sim_data(n = 100, p = 200, q = 5, noise_weight = 0.5)
```

```{r, cache=TRUE}
sim7_pls <- ddsPLS(sim_7$X, sim_7$Y,
                   verbose = TRUE)
```
```{r, cache=TRUE}
set.seed(12)

sim_8 <- sim_data(n = 100, p = 200, q = 150, noise_weight = 0)
```

```{r, cache=TRUE}
sim8_pls <- ddsPLS(sim_8$X, sim_8$Y,
                   verbose = TRUE)
```
If $q > n$, then the model is still able to perfectly explain the variance.


```{r, cache=TRUE}
set.seed(12)

sim_9 <- sim_data(n = 50, p = 50, q = 150, noise_weight = 0)
```

```{r, cache=TRUE}
sim9_pls <- ddsPLS(sim_9$X, sim_9$Y,
                   verbose = TRUE)
```
Still works when $n<q$ and $p<q$.
